---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ragg

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/thomasp85/ragg.svg?branch=master)](https://travis-ci.org/thomasp85/ragg)
[![CRAN status](https://www.r-pkg.org/badges/version/ragg)](https://cran.r-project.org/package=ragg)
<!-- badges: end -->

This package provides graphic devices for R based on the AGG library developed
by the late Maxim Shemanarev. AGG provides both higher performance and higher
quality than the standard cairo based drivers provided by grDevices, and is
further system agnostic so that output should be identical across operating
systems.

## Installation
The package is currently under development and has only been configured for
Mac and Linux. You'll need to have pkg-config, freetype, fontconfig, libpng, 
and libtiff available on your system to compile the package. If you have these
libraries available you should be able to install the package using devtools:

```{r, eval=FALSE}
# install.packages('devtools')
devtools::install_github('thomasp85/ragg')
```

