---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ragg

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/thomasp85/ragg.svg?branch=master)](https://travis-ci.org/thomasp85/ragg)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/thomasp85/ragg?branch=master&svg=true)](https://ci.appveyor.com/project/thomasp85/ragg)
[![CRAN status](https://www.r-pkg.org/badges/version/ragg)](https://cran.r-project.org/package=ragg)
<!-- badges: end -->

This package provides graphic devices for R based on the AGG library developed
by the late Maxim Shemanarev. AGG provides both higher performance and higher
quality than the standard cairo based drivers provided by grDevices, and is
further system agnostic so that output should be identical across operating
systems.

## Installation
The package is currently under development and has only been configured for
Mac and Linux. You'll need to have pkg-config, freetype, fontconfig, libpng, 
and libtiff available on your system to compile the package. If you have these
libraries available you should be able to install the package using devtools:

```{r, eval=FALSE}
# install.packages('devtools')
devtools::install_github('thomasp85/ragg')
```

## Use
ragg provides _almost_ drop-in replacements for the png and tiff graphic devices
provided by default from the grDevices packages and can both produce png and 
tiff files. Notable features, that sets itself apart from the build-in
devices, includes:

- Faster (up to 40% faster than anti-aliased cairo device)
- Direct access to all system fonts
- High quality anti-aliasing
- High quality rotated text
- Support 16-bit output
- System independent rendering (output from Mac, Windows, and Linux should be
  identical)

You can use it like any other device. The main functions are `agg_png()` and
`agg_tiff()`, both have arguments that closely match that of the `png()` and 
`tiff()` functions, so switching over should be easy.

```{r}
library(ragg)
library(ggplot2)

file <- knitr::fig_path('.png')

agg_png(file, width = 1000, height = 500, res = 144)
ggplot(mtcars) + 
  geom_point(aes(mpg, disp, colour = hp)) + 
  labs(title = 'System fonts — Oh My!') + 
  theme(text = element_text(family = 'Papyrus'))
invisible(dev.off())

knitr::include_graphics(file)
```

