---
title: "Ragg Quality"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ragg Quality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7
)
```

```{r setup}
library(ragg)
library(grid)
library(magick)
has_xlib <- Sys.info()['sysname'] != 'Windows'
img_zoom <- function(path) {
  img <- image_read(path)
  img <- image_crop(img, '300x150+150+75')
  img <- image_sample(img, '600x300')
  img
}
```

This vignette tries to compare the quality of the  output of ragg, with that of
cairo and, if the system supports it, Xlib (Xlib is unix only). As quality is, 
to a certain extend, subjective, the vignette will be short on conclusions and
filled with examples.

## Scope
There are mainly two areas of high importance when discussing graphic device 
quality: Shape rendering (fill and stroke) and text rendering. Both of these are
highly dependent on the quality of anti-aliasing if any. Apart from that we will 
also look into alpha blending and raster interpolation.

## Shape rendering
The examples here are relevant for rendering of all different types of shapes, 
be it lines or polygons. Lines obviously doesn't have fill, so only the stroke
rendering will have relevance here. We chose to use a circle for this as it 
provides a nice sampling of uneven edges.

```{r}
circle_quality <- function(device, name, file, ...) {
  device(file, width = 600, height = 300, ...)
  grid.circle(
    x = c(0.25, 0.75), 
    r = 0.4, 
    gp = gpar(col = c('black', NA), fill = c(NA, 'black'), lwd = 2)
  )
  grid.text(y = 0.1, label = name, gp = gpar(cex = 2))
  invisible(dev.off())
}
```

```{r}
ragg_circle <- knitr::fig_path('.png')

circle_quality(agg_png, 'ragg', ragg_circle)

knitr::include_graphics(ragg_circle)
```

```{r, echo=FALSE}
img_zoom(ragg_circle)
```

```{r, eval=has_xlib, include=has_xlib}
xlib_circle <- knitr::fig_path('.png')

circle_quality(png, 'Xlib', xlib_circle, type = 'Xlib')

knitr::include_graphics(xlib_circle)
```

```{r, eval=has_xlib, include=has_xlib, echo=FALSE}
img_zoom(xlib_circle)
```

```{r}
cairo_none_circle <- knitr::fig_path('.png')

circle_quality(png, 'cairo no AA', cairo_none_circle, 
               type = 'cairo', antialias = 'none')

knitr::include_graphics(cairo_none_circle)
```

```{r, echo=FALSE}
img_zoom(cairo_none_circle)
```

```{r}
cairo_gray_circle <- knitr::fig_path('.png')

circle_quality(png, 'cairo gray AA', cairo_gray_circle, 
               type = 'cairo', antialias = 'gray')

knitr::include_graphics(cairo_gray_circle)
```

```{r, echo=FALSE}
img_zoom(cairo_gray_circle)
```

```{r}
cairo_subpixel_circle <- knitr::fig_path('.png')

circle_quality(png, 'cairo subpixel AA', cairo_subpixel_circle, 
               type = 'cairo', antialias = 'subpixel')

knitr::include_graphics(cairo_subpixel_circle)
```

```{r, echo=FALSE}
img_zoom(cairo_subpixel_circle)
```

### Observations
ragg is the only device that provides anti-aliasing of fill, which results in
obvious quality differences. The reason for not doing so in cairo is presumably
to avoid artefacts when shapes are touching each other where anti-aliasing can
result in a thin edge between the two shapes instead of a contiguous colour. 
This is a real issue, but I personally don't agree that it should be allowed to
degrade the overall quality of the device, hence the reason for not following
this approach in ragg.

Xlib (if that is available on your system), provides completely non anti-aliased
output and so does cairo with  `antialias = 'none'`. It is surprising that Xlib
appears to have a better stroke rendering than cairo without anti-aliasing. 
Further, the difference between `'gray'` and `'subpixel'` antialiasing is not
visible to the naked eye, nor with a 2x zoom.

## Text
Text is a difficult thing to handle for a graphic device, both in terms of 
finding system fonts, and in terms of rendering. Often the rendering is 
offloaded to another library (e.g. freetype), which will provide a bitmap 
representation to blend into the device. This approach is often good for 
horizontal or vertical text, but struggle with other rotations. Here we will 
test text rendering at different sizes and at a 27&#176; counter-clockwise 
rotation. We will use the the system provided *serif* font as it provides a more
complex task than a *sans-serif* one.

```{r}
text_quality <- function(device, name, file, rotation = 0, ...) {
  text <- 'The quick blue R jumped over the lazy snake'
  vp <- viewport(angle = rotation)
  device(file, width = 600, height = 300, ...)
  pushViewport(vp)
  grid.text(text, x = 0.1, y = 0.2, just = 'left', gp = gpar(fontfamily = 'serif', cex = 0.5))
  grid.text(text, x = 0.1, y = 0.4, just = 'left', gp = gpar(fontfamily = 'serif', cex = 1))
  grid.text(text, x = 0.1, y = 0.6, just = 'left', gp = gpar(fontfamily = 'serif', cex = 1.5))
  grid.text(text, x = 0.1, y = 0.8, just = 'left', gp = gpar(fontfamily = 'serif', cex = 2))
  popViewport()
  grid.text(x = 0.9, y = 0.1, label = name, just = 'right', gp = gpar(cex = 2))
  invisible(dev.off())
}
```

```{r}
ragg_text <- knitr::fig_path('.png')

text_quality(agg_png, 'ragg', ragg_text)

knitr::include_graphics(ragg_text)
```

```{r, echo=FALSE}
img_zoom(ragg_text)
```

```{r}
ragg_text_rot <- knitr::fig_path('.png')

text_quality(agg_png, 'ragg', ragg_text_rot, rotation = 27)

knitr::include_graphics(ragg_text_rot)
```

```{r, echo=FALSE}
img_zoom(ragg_text_rot)
```

```{r, eval=has_xlib, include=has_xlib}
xlib_text <- knitr::fig_path('.png')

text_quality(png, 'Xlib', xlib_text, type = 'Xlib')

knitr::include_graphics(xlib_text)
```

```{r, eval=has_xlib, include=has_xlib, echo=FALSE}
img_zoom(xlib_text)
```

```{r, eval=has_xlib, include=has_xlib}
xlib_text_rot <- knitr::fig_path('.png')

text_quality(png, 'Xlib', xlib_text_rot, rotation = 27, type = 'Xlib')

knitr::include_graphics(xlib_text_rot)
```

```{r, eval=has_xlib, include=has_xlib, echo=FALSE}
img_zoom(xlib_text_rot)
```

```{r}
cairo_none_text <- knitr::fig_path('.png')

text_quality(png, 'cairo no AA', cairo_none_text, 
             type = 'cairo', antialias = 'none')

knitr::include_graphics(cairo_none_text)
```

```{r, echo=FALSE}
img_zoom(cairo_none_text)
```

```{r}
cairo_none_text_rot <- knitr::fig_path('.png')

text_quality(png, 'cairo no AA', cairo_none_text_rot, rotation = 27, 
             type = 'cairo', antialias = 'none')

knitr::include_graphics(cairo_none_text_rot)
```

```{r, echo=FALSE}
img_zoom(cairo_none_text_rot)
```

```{r}
cairo_gray_text <- knitr::fig_path('.png')

text_quality(png, 'cairo gray AA', cairo_gray_text, 
             type = 'cairo', antialias = 'gray')

knitr::include_graphics(cairo_gray_text)
```

```{r, echo=FALSE}
img_zoom(cairo_gray_text)
```

```{r}
cairo_gray_text_rot <- knitr::fig_path('.png')

text_quality(png, 'cairo gray AA', cairo_gray_text_rot, rotation = 27, 
             type = 'cairo', antialias = 'gray')

knitr::include_graphics(cairo_gray_text_rot)
```

```{r, echo=FALSE}
img_zoom(cairo_gray_text_rot)
```

```{r}
cairo_subpixel_text <- knitr::fig_path('.png')

text_quality(png, 'cairo subpixel AA', cairo_subpixel_text, 
             type = 'cairo', antialias = 'subpixel')

knitr::include_graphics(cairo_subpixel_text)
```

```{r, echo=FALSE}
img_zoom(cairo_subpixel_text)
```

```{r}
cairo_subpixel_text_rot <- knitr::fig_path('.png')

text_quality(png, 'cairo subpixel AA', cairo_subpixel_text_rot, rotation = 27, 
             type = 'cairo', antialias = 'subpixel')

knitr::include_graphics(cairo_subpixel_text_rot)
```

```{r, echo=FALSE}
img_zoom(cairo_subpixel_text_rot)
```

### Observations
Font handling is hard... Setting the font to `'serif'` means different things to
different devices and being more specific results in some devices not being able
to find the font. ragg exclusively uses the system fonts so whatever your OS 
defines as the base serif type it will pick up. Cairo goes through the internal
R database to pick a slightly different font. Xlib... well I can't comment on 
what it is doing, but it appears to pick something completely different from the
X11 system. Regarding quality, I don't think I have to be mean and mention Xlib,
so let's look at ragg and cairo. Ignoring for a fact that they have used two 
different fonts, we can see some differences and some interesting stuff. First,
cairo don't care about the anti-alias setting when it renders fonts. My guess is
that it will always ask for an 8bit pixelmap from the font engine (probably 
freetype) and use that. This ensures high quality fonts no matter the settings.
The text appears quite heavy, which (unless this is a feature of the font) 
indicates that cairo does not gamma-correct the pixelmap before blending it into
the image. Correct gamma correction of font is quite important, so if that is 
the case it is quite sad. Another thing we notice is that cairo uses the 
pixelmaps even for rotated text. This result in jagged baseline and uneven 
kerning when plotting rotated text. ragg only uses pixelmaps when plotting 
axis-aligned text. For rotated text it will convert the glyphs to polygons and
render them using the built-in rasterizer ensuring an even baseline and kerning.
This means plotting rotated text is slightly less performant, at the cost of 
looking good — I can live with that trade-off.

## Alpha blending
How transparent colours are combined is not necessarily equal among devices. The
biggest divide is on whether to use premultiplied colours or not. With 
premultiplied colours the red, green, and blue intensity is wieghted by the 
alpha directly, instead of alpha simply being an additional value. Using 
premultiplied colours is the only way to get correct alpha blending. As the Xlib
device doesn't support transparent colours it will be exempt from this 
comparison even on systems that have it. Further, as alpha blending is not 
related to anti-aliasing, we will simply compare ragg against a single cairo
setup.

```{r}
blend_quality <- function(device, name, file, ...) {
  device(file, width = 600, height = 300, ...)
  grid.rect(x = 0.35, y = 0.4, width = 0.5, height = 0.5, gp = gpar(fill = '#FF0000'))
  grid.rect(x = 0.65, y = 0.6, width = 0.5, height = 0.5, gp = gpar(fill = '#00FF001A'))
  grid.text(x = 0.9, y = 0.1, label = name, just = 'right', gp = gpar(cex = 2))
  invisible(dev.off())
}
```

```{r}
ragg_blend <- knitr::fig_path('.png')

blend_quality(agg_png, 'ragg', ragg_blend)

knitr::include_graphics(ragg_blend)
```

```{r}
cairo_blend <- knitr::fig_path('.png')

blend_quality(png, 'cairo', cairo_blend, type = 'cairo')

knitr::include_graphics(cairo_blend)
```

### Observations
Nothing much to say — both devices handles alpha blending correctly (see  [here](https://developer.nvidia.com/content/alpha-blending-pre-or-not-pre) to
understand what this test was about)

## Raster
The main way raster plotting can get influence by the device is in how the image
gets interpolated during scaling, which will be briefly compared here.

```{r}
raster_quality <- function(device, name, file, ...) {
  reds <- matrix(hcl(0, 80, seq(50, 80, 10)),
                        nrow = 4, ncol = 5)
  device(file, width = 600, height = 300, ...)
  grid.raster(reds, vp = viewport(0.25, 0.25, 0.5, 0.5, angle = 27))
  grid.raster(reds, interpolate = FALSE, 
              vp = viewport(0.75, 0.75, 0.5, 0.5, angle = 27))
  grid.text(x = 0.9, y = 0.1, label = name, just = 'right', gp = gpar(cex = 2))
  invisible(dev.off())
}
```

```{r}
ragg_raster <- knitr::fig_path('.png')

raster_quality(agg_png, 'ragg', ragg_raster)

knitr::include_graphics(ragg_raster)
```

```{r, eval=has_xlib, include=has_xlib}
xlib_raster <- knitr::fig_path('.png')

raster_quality(png, 'Xlib', xlib_raster, type = 'Xlib')

knitr::include_graphics(xlib_raster)
```

```{r}
cairo_raster <- knitr::fig_path('.png')

raster_quality(png, 'cairo', cairo_raster, type = 'cairo')

knitr::include_graphics(cairo_raster)
```

### Observations
As with blending there is nothing much to see here. All devices perform equally
and correctly.

## Conclusion
When it comes to raster quality there is only really the two contenders are 
obviously ragg and anti-aliased cairo, as lack of anti-aliasing has clear 
detrimental effect on output quality. As there appear to be no real difference 
in quality between cairo's two anti-aliasing modes the question basically boils
down to cairo vs ragg. While for the most part the two rendering systems provide
comparable output, there are two areas where ragg takes the lead, quality-wise:
rendering of fill, and rendering of rotated text. If these areas are of interest 
to you then ragg will be the obvious choice.
